Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-08-14 analysis/paper.tex
Update results and discussion

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. The models performance in
predicting the occurrence of specific species at archaeological sites is
highly variable, but in aggregate the predictions are coherent and align
with broad-scale trends in the ubiquity and relative abundance of
species in the archaeological record.
\end{abstract}

@@

Biogeographic research into the centres of origin and/or domestication
of crops has also long informed broader understanding of the process of
agricultural origins. In \emph{Man Makes Himself}, \citet{Childe1936}
correctly located the centre of origin of European agriculture in the
`Fertile Crescent' of West Asia (unlike for example \citet{Pumpelly1908}
before him). This was not based on the region's prehistoric
archaeological record, which at that point was barely known. Instead he
was guided to the region by biogeographic work by Vavilov and
\citet{PeakeFleure1927}; only later was this prediction validated by
archaeological work on the Epipalaeolithic and Neolithic of Palestine
\citep{Boyd2018}. In subsequent decades, the search for more precise
origin zones of specific domestic plants relied on the assumption that
``the locus of domestication of a wild plant would presumably be within
its area of original distribution in the wild state''
\citetext{\citealp{Butzer1971}; \citealp[paraphrasing][]{Helbaek1959}}
-- and that this ``natural habitat'' has not changed significantly over
the last 12,000 years \citep{Butzer1971}.

@@

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the range of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---or more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguihed from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Based on current understandings, we
assume that plant-based subsistence during this period was broad,
geographically- and temporally-varied, and reflects a gradual,
geographically decentred, and nonlinear tranistion to greater reliance
on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
point was a list of 97 taxa (Table~\ref{tbl-results-summary}) comprising
the identifiable species observed at at least two Late
Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset collated from
three previously published regional archaeobotanical databases: ADEMNES
\citep{ADEMNES}, ORIGINS \citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

The taxonomic identifications of archaeobotanical material given in our
source databases were controlled to ensure consistency between sources
and to group taxa that cannot always be reliably distinguished
\citep[for details see][]{ArranzOtaeguiRoe2023}. Taxonomic names were
then matched to the canonical form specified in the GBIF Backbone
Taxonomy \citep{GBIFSecretariat2023} so they could be related to modern
occurrences. The aim in both cases was to resolve taxonomic ambiguity by
using a more inclusive taxonomic unit. For example, the model discussed
here as \emph{Bolboschoenus} includes both archaeological occurrences of
\emph{B. maritimus} and \emph{B. glaucus}, since these were not
distinguished by archaeobotanists until recently
\citep{WollstonecroftEtAl2011}, as well as modern occurrences under the
older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where
different.

@@

Georeferenced occurrence data for West Asia was obtained from the Global
Biodiversity Information Facility (GBIF) using via its application
programming interface and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. GBIF was cleaned to removed
fossil occurrences, recorded absences, and records with missing,
imprecise (\textgreater1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small
training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
excluded 14 taxa with less than 50 usable occurrences, following
recommendations for niche models generally and Random Forest-based
models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
as this would have been computationally prohibitive and we were
uncertain what accounted for such a disproportionately high number of
records.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia used to train models (N=81641}

@@

GBIF is currently the best available general-purpose occurrence dataset
for the West Asia region, its coverage is uneven both geographically and
from species to species. Figure~\ref{fig-occ-map} shows clearly that the
Southern Levant, and Israel specifically, is significantly more densely
sampled than other regions.

Random Forest is a presence--absence approach to niche modelling and
therefore requires not just data on where a species is present, but
where it is definitely not present. However `absence' data is rarely
available because it requires exhaustive survey. In practice, most
applications of niche modelling are `presence-only' and, where absence
data is required (as for Random Forest), it is supplied as a random
background sample of `pseudo-absence' points. The purpose of this sample
is to inform the model about the nature of the underlying environment.
The stochastic generation process means that some `pseudo-absence'
points will overlap or fall close to presences, so ensuring the model is
not overly influenced by background samples is critical to its
predictive importance \citep{ValaviEtAl2022}. Here we follow the advice
of \citet{BarbetMassinEtAl2012} for regression-based species
distribution models and use a large (≈10000) uniform sample of points
from across the land area of the study region. These points are then
weighted equally against the presences in the regression to produce a
`balanced Random Forest' \citep{ValaviEtAl2022}.

@@

@@

\begin{itemize}
\tightlist
\item
  Terrain aspect and slope, which at high resolution perform well as
  proxies for solar radiation when modelling plant occurrence
  \citep{AustinVanNiel2011, LeempoelEtAl2015}; and the topographic
  wetness index (TWI), which serves as a proxy for soil moisture and is
  particularly important in modelling arid environments
  \citep{KopeckyCizkova2010, CamposEtAl2016, DiVirgilioEtAl2018}. All
  three were derived from the SRTM15+ digital elevation model using
  algorithms from WhiteboxTools \citep{Lindsay2016}.
\end{itemize}

@@

@@

We trained Random Forest models for 82 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Section~\ref{sec-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 328 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition
Table~\ref{tbl-results-summary}. Model accuracy (proportion of correctly
classified presence and background samples) ranged between 73\% and
99\%, with an average of 93\%. Sensitivity (proportion of correctly
classified presence samples) ranged between 79\% and 100\%, with an
average of 94\%. The area under the models' receiver operating
characteristic curves (ROC-AUC) was on average 0.978±0.061 Model
sensitivity is loosely correlated with the number of occurrences
available for training (Figure~\ref{fig-model-sensitivity-vs-n_occ}),
with the worst-performing models all having less than 300 recorded
occurrences: \emph{Asphodelus aestivus}, \emph{Triticum turgidum
dicoccum}, \emph{Zygophyllum fabago}, and \emph{Stipa dregeana}. Test
metrics and ROC curves for the individual models are included in the
appendix.

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is worse, with 10\% of presences in
archaeobotanical assemblages successfully predicted. Model sensitivity
(proportion of corrected predicted presenses) in relation to the
archaeological data is on average 0.06±0.13
Table~\ref{tbl-results-summary}. A full assessment of the hindcasting
performance of the individual models can be found in the appendix.

@@

{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.86 & 0.00 \\ 
{\itshape Aegilops speltoides} & — & 701 & 0.92 & 0.98 & 0.95 & — \\ 
{\itshape Aegilops tauschii} & — & 725 & 0.89 & 0.98 & 0.94 & — \\ 
{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 515 & 0.98 & 1.00 & 0.96 & 0.00 \\ 
{\itshape Ammi majus} & 4 & 900 & 0.98 & 0.99 & 0.97 & 0.00 \\ 
{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.95 & 0.88 & 0.12 \\ 
{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.84 & 0.95 & 0.91 & 0.00 \\ 
{\itshape Arnebia decumbens} & 21 & 324 & 0.95 & 0.95 & 0.89 & 0.00 \\ 
{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 0.99 & 0.94 & 0.00 \\ 
{\itshape Asphodelus aestivus} & 2 & 76 & 0.97 & 0.97 & 0.83 & 0.00 \\ 
{\itshape Atriplex prostrata} & 4 & 266 & 0.98 & 1.00 & 0.94 & 0.00 \\ 
{\itshape Avena fatua} & 2 & 579 & 0.86 & 0.93 & 0.86 & 0.00 \\ 
{\itshape Avena sterilis} & 5 & 53244\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
{\itshape Bassia prostrata} & 2 & 46\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Bolboschoenus glaucus} & 5 & 27\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{3}} & 31 & 1040 & 0.98 & 0.99 & 0.97 & 0.03 \\ 
{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 & 0.00 \\ 
{\itshape Bromus sterilis} & 3 & 759 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
{\itshape Buglossoides arvensis} & 23 & 260 & 0.87 & 0.97 & 0.95 & 0.09 \\ 
{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.95 & 0.08 \\ 
{\itshape Camphorosma monspeliaca} & 2 & 38\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 & 0.00 \\ 
{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.94 & 0.00 \\ 
{\itshape Celtis tournefortii} & 2 & 14\textsuperscript{\textit{2}} & — & — & — & — \\ 
{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 & 0.50 \\ 
{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.94 & 0.00 \\ 

>>>>

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.

The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. The models performance in
predicting the occurrence of specific species at archaeological sites is
highly variable, but in aggregate the predictions are coherent and align
with broad-scale trends in the ubiquity and relative abundance of
species in the archaeological record.
\end{abstract}

@@

Biogeographic research into the centres of origin and/or domestication
of crops has also long informed broader understanding of the process of
agricultural origins. In \emph{Man Makes Himself}, \citet{Childe1936}
correctly located the centre of origin of European agriculture in the
`Fertile Crescent' of West Asia (unlike for example \citet{Pumpelly1908}
before him). This was not based on the region's prehistoric
archaeological record, which at the time had only been cursorily
explored. Instead he was guided to the region by biogeographic work by
Vavilov and \citet{PeakeFleure1927}; only later was this prediction
validated by archaeological work on the Epipalaeolithic and Neolithic of
Palestine \citep{Boyd2018}. In subsequent decades, the search for more
precise origin zones of specific domestic plants relied on the
assumption that ``the locus of domestication of a wild plant would
presumably be within its area of original distribution in the wild
state''
\citetext{\citealp{Butzer1971}; \citealp[paraphrasing][]{Helbaek1959}}
-- and that this ``natural habitat'' has not changed significantly over
the last 12,000 years \citep{Butzer1971}.

@@

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the range of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---or more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguished from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Based on current understandings, we
assume that plant-based subsistence during this period was broad,
geographically- and temporally-varied, and reflects a gradual,
geographically decentred, and nonlinear tranistion to greater reliance
on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
point was a list of 66 taxa (Table~\ref{tbl-results-summary}) comprising
the identified species observed at at least 3 Late
Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset collated from
three previously published regional archaeobotanical databases: ADEMNES
\citep{ADEMNES}, ORIGINS \citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

The taxonomic identifications of archaeobotanical material given in our
source databases were previously controlled to ensure consistency
between sources and to remove taxa that cannot be reliably distinguished
\citep[for details see][]{ArranzOtaeguiRoe2023}. Taxonomic names were
then matched to the canonical form specified in the GBIF Backbone
Taxonomy \citep{GBIFSecretariat2023} so they could be related to modern
occurrences. Archaeologically-attested domestic species meeting our
inclusion criteria were substituted for their wild progenitors (where
different) when gathering occurrence data, since the domestic forms are
now widespread and presumably uninformative of the species original
niche.

@@

\begin{verbatim}
Rows: 1 Columns: 11
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (5): key, doi, license, status, downloadLink
dbl  (3): size, totalRecords, numberDatasets
dttm (3): created, modified, eraseAfter

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

Georeferenced occurrence data for West Eurasia between 0 and 60° of
latitude was obtained from the Global Biodiversity Information Facility
(GBIF) using via and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. The GBIF dataset
\citep{GBIFDerivedData} excluded fossil occurrences, recorded absences,
and records with missing or dubious coordinates. Although niche models
have reasonable predictive power even with small training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we did
not attempt to model 3 taxa with less than 40 usable occurrences,
following recommendations for niche models generally and Random
Forest-based models specifically
\citep{StockwellPeterson2002, LuanEtAl2020}. Multiple records of the
same taxon at the same coordinate were discarded because they do not
impart information to the model. The resulting cleaned dataset used to
train our niche models comprises \ensuremath{3.387196\times 10^{6}} from
4764 constituent datasets.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia used to train models (N=1409939}

@@

GBIF is currently the best available general-purpose occurrence dataset
for the West Asia region, its coverage is uneven both geographically and
from species to species. Figure~\ref{fig-occ-map} shows clearly that the
Southern Levant, and Israel specifically, is significantly more densely
sampled than other parts of West Asia.

Random Forest is a presence--absence approach to niche modelling and
therefore requires not just data on where a species is present, but
where it is definitely not present. However `absence' data is rarely
available because it requires exhaustive survey. In practice, most
applications of niche modelling are `presence-only' and, where absence
data is required (as for Random Forest), it is supplied as a random
background sample of points.\footnote{We use a background sample that is
  representative of the entire study area, as opposed to attempting to
  construct a `pseudo-absence' sample. See \citet{SilleroEtAl2021} for
  the distinction.} The purpose of this sample is to inform the model
about the nature of the underlying environment. The stochastic
generation process means that some of these points will overlap or fall
close to presences, so ensuring the model is not overly influenced by
background samples is critical to its predictive importance
\citep{ValaviEtAl2022}. Here we follow the advice of
\citet{BarbetMassinEtAl2012} for regression-based species distribution
models and use a large (≈10000) uniform sample of points from across the
land area of the study region. These points are then weighted equally
against the presences in the regression to produce a `balanced Random
Forest' \citep{ValaviEtAl2022}.

@@

\begin{verbatim}
bio_1, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
\end{verbatim}

\begin{verbatim}
Warning in parse_netcdf_meta(meta_data, get_names(x)): NAs introduced by
coercion
\end{verbatim}

\begin{verbatim}
bio_10, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_11, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_12, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_13, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_14, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_15, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_16, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_17, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_18, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_19, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_2, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_3, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_4, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_5, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_6, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_7, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_8, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_9, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

@@

\begin{itemize}
\tightlist
\item
  Terrain aspect and slope, which at high resolution perform well as
  proxies for solar radiation when modelling plant occurrence
  \citep{AustinVanNiel2011, LeempoelEtAl2015}; and the topographic
  wetness index (TWI), which serves as a proxy for soil moisture and is
  particularly important in modelling arid environments
  \citep{KopeckyCizkova2010, CamposEtAl2016, DiVirgilioEtAl2018}. All
  three were derived from the SRTM30+ digital elevation model using
  algorithms from WhiteboxTools \citep{Lindsay2016}.
\end{itemize}

@@

\begin{verbatim}
Warning: There were 57 warnings in `mutate()`.
The first warning was:
i In argument: `occ = map(occ, bind_extract, st_as_stars(hack(climate,
  "period", "cur")))`.
Caused by warning in `CPL_get_metadata()`:
! GDAL Message 1: dimension #0 (period) is not a Time or Vertical dimension.
i Run `dplyr::last_dplyr_warnings()` to see the 56 remaining warnings.
\end{verbatim}

@@

We trained Random Forest models for 63 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Section~\ref{sec-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 252 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition
Table~\ref{tbl-results-summary}. Model accuracy (proportion of correctly
classified presence and background samples) ranged between 92\% and
100\%, with an average of 97\%. Sensitivity (proportion of correctly
classified presence samples) ranged between 12\% and 100\%, with an
average of 82\%. The area under the models' receiver operating
characteristic curves (ROC-AUC) was on average 0.987±0.018 Model
sensitivity is loosely correlated with the number of occurrences
available for training (Figure~\ref{fig-model-sensitivity-vs-n_occ}),
with the worst-performing models all having less than 300 recorded
occurrences: \emph{Lathyrus sativus}, \emph{Lepidium perfoliatum},
\emph{Triticum aestivum}, \emph{Bolboschoenus glaucus}, \emph{Gypsophila
vaccaria}, \emph{Euclidium syriacum}, \emph{Gypsophila elegans},
\emph{Citrullus colocynthis}, \emph{Triticum monococcum}, \emph{Triticum
turgidum}, \emph{Aegilops crassa}, \emph{Secale strictum}, \emph{Cicer
reticulatum}, \emph{Medicago astroites}, \emph{Triticum durum},
\emph{Gypsophila pilosa}, \emph{Arnebia decumbens}, \emph{Salvia
absconditiflora}, \emph{Suaeda fruticosa}, \emph{Buglossoides
tenuiflora}, \emph{Lathyrus oleraceus}, \emph{Vitis sylvestris}, and
\emph{Vicia orientalis}. Test metrics and ROC curves for the individual
models are included in the appendix.

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is worse, with 10\% of presences in
archaeobotanical assemblages successfully predicted. Model sensitivity
(proportion of corrected predicted presenses) in relation to the
archaeological data is on average 0.07±0.15
Table~\ref{tbl-results-summary}. A full assessment of the hindcasting
performance of the individual models can be found in the appendix.

@@

{\itshape Aegilops crassa} & 4 & 223 & 0.98 & 0.98 & 0.43 & 0.00 \\ 
{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 762 & 0.99 & 1.00 & 0.88 & 0.50 \\ 
{\itshape Ammi majus} & 4 & 10311 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
{\itshape Androsace maxima} & 17 & 4946 & 0.94 & 0.98 & 0.91 & 0.06 \\ 
{\itshape Arenaria serpyllifolia} & 3 & 101612 & 0.98 & 0.98 & 1.00 & 0.00 \\ 
{\itshape Arnebia decumbens} & 21 & 616 & 0.97 & 0.97 & 0.59 & 0.05 \\ 
{\itshape Arnebia linearifolia} & 13 & 239 & 1.00 & 0.99 & 0.89 & 0.00 \\ 
{\itshape Atriplex prostrata} & 4 & 80181 & 0.98 & 0.99 & 1.00 & 0.00 \\ 
{\itshape Avena sterilis} & 5 & 15906 & 0.96 & 0.99 & 0.99 & 0.40 \\ 
{\itshape Bassia arabica} & 4 & 262 & 1.00 & 1.00 & 0.95 & 0.00 \\ 
{\itshape Bolboschoenus glaucus} & 5 & 418 & 0.98 & 0.97 & 0.53 & 0.00 \\ 
{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{2}} & 31 & 53809 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
{\itshape Brachypodium distachyon} & 5 & 13800 & 0.98 & 1.00 & 0.99 & 0.20 \\ 
{\itshape Bromus sterilis} & 3 & 89764 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
{\itshape Buglossoides arvensis} & 23 & 23004 & 0.95 & 0.98 & 0.98 & 0.09 \\ 
{\itshape Buglossoides tenuiflora} & 26 & 399 & 0.99 & 0.99 & 0.71 & 0.04 \\ 
{\itshape Capparis spinosa} & 4 & 8617 & 0.97 & 0.99 & 0.96 & 0.00 \\ 
{\itshape Carex divisa} & 9 & 7400 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
{\itshape Chenopodium album} & 5 & 184403 & 0.99 & 0.98 & 1.00 & 0.00 \\ 
{\itshape Cicer reticulatum}\textsuperscript{\textit{3}} & 3 & 52 & 1.00 & 1.00 & 0.69 & 0.00 \\ 
{\itshape Citrullus colocynthis} & 4 & 1823 & 0.94 & 0.95 & 0.69 & 0.00 \\ 
{\itshape Crithopsis delileana} & 3 & 430 & 1.00 & 1.00 & 0.93 & 0.00 \\ 
{\itshape Euclidium syriacum} & 3 & 338 & 0.98 & 0.97 & 0.27 & 0.00 \\ 
{\itshape Ficus carica} & 8 & 64424 & 0.98 & 0.99 & 0.99 & 0.38 \\ 
{\itshape Fumaria densiflora} & 4 & 3704 & 0.96 & 0.99 & 0.94 & 0.00 \\ 
{\itshape Gypsophila elegans} & 3 & 1603 & 0.95 & 0.98 & 0.81 & 0.00 \\ 
{\itshape Gypsophila pilosa} & 5 & 348 & 0.99 & 0.99 & 0.67 & 0.00 \\ 
{\itshape Gypsophila vaccaria}\textsuperscript{\textit{4}} & 8 & 1177 & 0.95 & 0.98 & 0.65 & 0.00 \\ 
{\itshape Henrardia pubescens} & 3 & 1\textsuperscript{\textit{5}} & — & — & — & — \\ 
{\itshape Hordeum bulbosum} & 4 & 4134 & 0.97 & 1.00 & 0.95 & 0.00 \\ 
